---
title: "Part 2 - Working with Factors"
author: "Ted Laderas"
date: "5/18/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## R Markdown

In this section, we'll learn some skills about manipulating factor (categorical) data.

We'll do this by making a bar plot and progressively making it more complex.

## Reviewing Factors 

Factors are how R represents categorical data.

There are two kinds of factors: 

+ `factor` - used for *nominal* data ("Ducks","Cats","Dogs")
+ `ordered`- used for *ordinal* data ("10-30","31-40","41-60")

We'll manipulate our barplot and add more information using factors.

```{r}
##make our table of pets
pets <- data.frame(id=c('C1','C2','D1','D4','D8', 'C4', 'D5', 'C10', "G4"),
                names=factor(c("Morris", "Morris", "Fido", "Mr Bowser", "Fido", "Lady Sheba", "Hubert", "Apples", "Nibbles")),
                weight=c(6,7,15,20,40, 10, 60, 17, 0.5),
                animal=factor(c("cat", "cat", "dog", "dog", "dog", "cat", "dog", "cat", "gerbil")),
                ##because it's an ordered category, we have to specify 
                ##order as the levels argument
                ageCategory= ordered(c("0-2", "5-8", "5-8", "3-4", "3-4", "5-8", "0-2", "5-8", "0-2"), 
                                       levels=c("0-2", "3-4", "5-8")),
                shotsCurrent=c("No", "Yes", NA, "No", "Yes", "No", "No", "Yes", NA))

pets
```

## `forcats` - a useful tidyverse package for doing stuff with categorical data

## Let's build a bar plot

Let's plot the weight for each pet.

```{r}
ggplot(data=pets, aes(x=id,y=weight)) + geom_bar(stat="identity")
```

## Sort by another variable

Let's sort by weight. We can do this by adding a `fct_reorder()` expression to define a new variable `id2` whose categories are ordered by `weight`.

```{r}
library(forcats)
pets %>% mutate(id2=fct_reorder(id, weight))  %>% ggplot(aes(x=id2,y=weight)) + geom_bar(stat="identity")
```

## What about counting?

The `geom_bar()` default is to count the number with each factor. Note that you don't map to a y-aesthetic here, because the y values are the counts.

Given this dataset, we want to ask how many pets have the same name.

Map another variable to `fill` (try both `weight` and `animal`). Are the results what you expected?

```{r}
ggplot(pets, aes(x=names)) + geom_bar()
```

## Changing the order of a factor

Sometimes you just don't like the default alphabetical order. We're going to do reverse alphabetical order.

```{r}
ggplot(pets, aes(x=names)) + geom_bar()

```

Sometimes you want a factor to be sorted by another variable. You can use `fct_reorder`.

## Faceting

Say you have another `factor` variable and you want to stratify the plots based on that. 
You can do that by supplying the name of that variable as a facet.

```{r}
ggplot(pets, aes(x=names)) + geom_bar() + facet_wrap(facets=~shotsCurrent)
```

You might notice that there are blank spots for the categories in each facet. We can restrict factors in each by using `scale="free"` argument in `facet_wrap()`.

```{r}
ggplot(pets, aes(x=names)) + geom_bar() + facet_wrap(facets=~shotsCurrent, scale="free")
```

## Stacked Bars

Let's see how many of each animal got shots. We can do this by mapping shots to `fill`.

```{r}
ggplot(pets, aes(x=animal,fill=shotsCurrent)) + geom_bar()

```

## Dodge those bars!

Instead of stacking, we can also dodge the bars (move the bars so they're beside each other).

```{r}
ggplot(pets, aes(x=animal,fill=shotsCurrent)) + geom_bar(position="dodge")

```

## Your Task

Given the `pets` `data.frame`, plot a stacked barchart showing that 

```{}

```

## Making an "other" category

Say we're only interested in young animals (`0-2`), and we want to lump the other age categories. 

We can use `fct_other`.

## Boxplots

Boxplots allow us to assess distributions of a continuous variable conditioned on categorical variables.

```{r}
ggplot(pets, aes(x=shotsCurrent, y=weight)) + geom_boxplot()

```

## Violin Plots

```{r}
ggplot(pets, aes(x=shotsCurrent, y=weight)) + geom_violin()

```

```{r}
data(iris)

ggplot(iris, aes(x=Species, y=Sepal.Length, fill=Species)) + geom_violin()
```
```
