---
title: "Part 2 - Working with Factors"
author: "Ted Laderas"
date: "5/18/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## R Markdown

In this section, we'll learn some skills about manipulating factor (categorical) data.

We'll do this by making a bar plot and a box plot and progressively making it more complex.

## Reviewing Factors 

Factors are how R represents categorical data.

There are two kinds of factors: 

+ `factor` - used for *nominal* data ("Ducks","Cats","Dogs")
+ `ordered`- used for *ordinal* data ("10-30","31-40","41-60")

We'll manipulate our barplot and add more information using factors.

```{r}
##make our table of pets
pets <- data.frame(id=c('C1','C2','D1','D4','D8', 'C4', 'D5', 'C10', "G4", "C11"),
                name=factor(c("Morris", "Morris", "Fido", "Mr Bowser", 
                              "Fido", "Lady Sheba", "Hubert", "Apples", "Nibbles",
                              "Morris")),
                ##weight is in pounds
                weight=c(6,7,15,20,40, 10, 60, 17, 0.5, 10),
                animal=factor(c("cat", "cat", "dog", "dog", "dog", 
                                "cat", "dog", "cat", "gerbil", "cat")),
                ##because it's an ordered category, we have to specify 
                ##order as the levels argument
                ageCategory= ordered(c("0-2", "5-8", "5-8", "3-4", "3-4", 
                                       "5-8", "0-2", "5-8", "0-2","5-8"), 
                                       levels=c("0-2", "3-4", "5-8")),
                shotsCurrent=c("No", "Yes", "No", "No", "Yes", "No", 
                               "No", "Yes", NA, "No"))

pets
```

## What about counting?

The `geom_bar()` default is to count the number of values with each factor level. Note that you don't map to a y-aesthetic here, because the y values are the counts.

Given this dataset, we want to ask how many pets have the same name. 

Try mapping another variable to `fill` (try both `weight` and `animal`). Are the results what you expected?

```{r}
ggplot(pets, aes(x=name)) + geom_bar()
```

## Faceting

Say you have another `factor` variable and you want to stratify the plots based on that. 
You can do that by supplying the name of that variable as a facet.

```{r}
ggplot(pets, aes(x=name)) + geom_bar() + 
  ## have to specify facets using notation
  ## try out facets=~ageCategory!
  facet_wrap(facets=~shotsCurrent) + 
  ## we make the x axis x angled for better legibility
  theme(axis.text.x = element_text(angle=45))
```

You might notice that there are blank spots for the categories in each facet. We can restrict factors in each by using `scale="free_x"` argument in `facet_wrap()`.

How many animals named "Morris" did not receive shots?

```{r}
ggplot(pets, aes(x=name)) + geom_bar() + 
  facet_wrap(facets=~shotsCurrent, scale="free_x") +
  theme(axis.text.x = element_text(angle=45))
  
```

## Stacked Bars

Let's see how many of each animal got shots. We can do this by mapping shots to `fill`.

```{r}
#we map color (the outline of the plot) to black to make it look prettier
ggplot(pets, aes(x=animal,fill=shotsCurrent)) + geom_bar(color="black")
```

## Proportional Barchart

We may only be interested in the relative proportions between the different categories. Visualizing this is useful for various 2 x 2 tests on proportions.

```{r}
ggplot(pets, aes(x=animal,fill=shotsCurrent)) + 
  geom_bar(position = "fill", color = "black")
```

## Dodge those bars!

Instead of stacking, we can also dodge the bars (move the bars so they're beside each other).

```{r}
ggplot(pets, aes(x=animal,fill=shotsCurrent)) + 
  geom_bar(position="dodge", color="black")

```

## Your Task

Given the `pets``data.frame`, plot a stacked proportional barchart showing that shows the age category counts by animal type. Is the proportion of animals receiving shots the same across each age category?

Intermediate: facet this plot by `shotsCurrent`. 

```{r}
pets %>% ggplot(aes(x=ageCategory, fill=animal)) + 
  geom_bar(position="fill", color="black") #+ 
  #facet_wrap(facets=~shotsCurrent, scales="free_x")
```

## Changing the order of a factor

Sometimes you just don't like the default alphabetical order. We're going to do reverse alphabetical order.

```{r}
ggplot(pets, aes(x=name)) + geom_bar()

```

Sometimes you want a factor to be sorted by another variable. You can use `fct_reorder`.

## Making an "other" category (intermediate)

Say we're only interested in comparing young animals (`0-2`) to the older animals. We can use `fct_other()`.

## Boxplots

Boxplots allow us to assess distributions of a continuous variable conditioned on categorical variables.

```{r}
ggplot(pets, aes(x=shotsCurrent, y=weight)) + geom_boxplot()
```

## Violin Plots

Violin plots are another useful way to visualize the data. Note that we only have one NA value, so we

```{r}
ggplot(pets, aes(x=shotsCurrent, y=weight)) + geom_violin()

```

## Looking at Sepal Length across Species

Our 

```{r}
data(iris)

ggplot(iris, aes(x=Species, y=Sepal.Length, fill=Species)) + geom_violin()
```

## More on manipulating factors

The following section is optional. Intermediate users may find these helpful.

You'll need to load the `forcats` package to use these.

## Let's build a bar plot

Let's plot the weight for each pet as a bar graph.

```{r}
ggplot(data=pets, aes(x=id,y=weight)) + geom_bar(stat="identity")
```

## Sort by another variable (intermediate)

Let's sort the barplot by weight. We can do this by adding a `fct_reorder()` expression to define a new variable `id2` whose categories are ordered by `weight`.

Based on this visualization, what can we conclude about the weights of each type of animal? Which kind of animal weighs the most?

```{r}
library(forcats)
pets %>% mutate(id2=fct_reorder(id, weight))%>% ggplot(aes(x=id2,y=weight)) + geom_bar(stat="identity")
```

## Plot in Reverse Alphabetical Order

Often, you want to plot things in reverse alphametical order. This is useful because heatmaps and such are often plotted from the bottom.

You can use `fct_rev` to do this.

```{r}
library(forcats)
pets %>% mutate(id2=fct_rev(id))%>% 
  ggplot(aes(x=id2,y=weight)) + 
  geom_bar(stat="identity")
```

## Sort by frequency

Going back to our `pets` data, sometimes we want to sort our count data by frequency. We can use `fct_infreq()` to do that.

```{r}
pets %>% mutate(name = fct_infreq(name)) %>% ggplot(aes(x=name)) + geom_bar()
```